Supp mat: time series paper

### 2017-06-13 writing/manuscript_draft.tex
moved results from ms to supp mat and expanded supp mat text

Addresses: \textbackslash{}
\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
\vspace{3 mm}

@@

Observational studies and population censuses have become a cornerstone
of modern ecology and conservation biology (Magurran et al. 2010).
Longterm data is necessary to understand population dynamics and to
assess extinction risk of species. Even though many time series may now
be considered ``longterm'', most are still short (i.e.~less than X
years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time. However,
there are of course notable exceptions including the Isle Royale study,
Iowa corn study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually
necessary? This question has important scientific and management
implications. Scientifically, we need to know how long a time series is
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when a trend in abundance over time is actually meaningful or
not. For example, the IUCN Red List Categories and Criteria suggest
under Criterian A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or 3 generations (IUCN 2012).
In addition, because sampling is typically expensive, we do not want to
sample for longer than is neccessary (Gerber, DeMaster, and Kareiva
1999).

@@

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (cite Bolker). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determinig sample
size, simulation approaches can be used (Gerrodette 1987). Simple
population models can be simulated with parameter values corresponding
to a population of interest. Out of a large number of simulations, power
can be calculated as the fraction of simulations that meet some
criteria. The specific criteria depend on the question at hand. Then
time series simulations of different lengths are run to determine the
minimum time series length required to be confident that you will be
able to answer your question (cite Bolker). In addition to using
simulated data, empirical time series can also be used. We usually do
not have multiple replicates of similar populations, but we can
subsample an empirical time series (cite). Subsamples of different
lengths can then be evaluated to see which fraction of subsamples meet
some criteria, again a measure of statistical power. Similar to the
simulation approach, we can use these measures of power to determine the
minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time. However, most
work has either focused on theoretical approaches or by examined a
single population. As an example of the former, Rhodes and Jonzen (2011)
examined the optimal allocation of effort between spatial and temporal
replicates. Using simple populations models, they found that the
allocation of effort depends on environmental variation, spatial and
temporal autocorrelation, and observer error. Rueda-Cediel et al. (2015)
also used a modeling approach, but parameterized it for a threatened
snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organims, that 15 years was adequate to assess longterm
trends in abundance. However, these studies, and other past work
(citation), have focused either on theoretical aspects of monitoring
design or focused on single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. To demonstrate these techniques, we address three
specific questions: (1) what is the minimum time series length required
to identify significant changes over time using linear regression, (2)

@@

In figure \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (t=40). It is clear
that statistical power increases quickly with increases in \(t\) (Fig.
\ref{fig:theoretical_approach}b). Once power is greater than XX (the
dotted line), that is the minimum time required (\(t_{min}\)

@@

Our objective is to determine the minimum time series length
(\(T_{min}\)) required to answer a specific question of interest. As we
describe below, the minimum time series length calculation is particular
to each question. Intuitively, we might expect a longer minimum time
series length (\(T_{min}\)) required given more complicated questions
(citations).

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, they are a representative sample. We first
take all possible contiguous subsamples of each time series. For
example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continueing until 1 possible contiguous subsample of length 40. In
this paper, we will try to ask a simple, specific question. We seek to
determine the \(T_{min}\) required to be confident in an estimate of the
slope coefficient from linear regression. Therefore, we run a linear
regression for each subsampled time series. Then, we determine the
fraction of subsamples of a particular length that have estimated slope
coefficients which are statistically different from zero given the
longterm, or ``true``, time series also has a significant slope. This is
a measure of statistical power (often denoted as \(1- \beta\), where
\(\beta\) is the probability of a type II error). Lastly, we determine
which subsample length is required to guarantee we have above a certain
threshold of statistical power, which we will assume is 0.8 by
convention (citation). A power of 0.8 would imply that given a trend
exists, there is a 0.8 probability of detecting the trend.
Complementary, there would be a 0.2 probability of a false negative.

@@

\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-3-1.pdf}

@@

We constructed a general population model where the trend over time
could be a model parameter. We chose a simple model, but any other
population model could be used. The specific model choice should be
tailored to the population of interest, based on its ecology. We then
ran simulations of different lengths to determine the minimum time
series length required to achieve a certain level of power (which we set
to 0.8). We were also able to alter the parameter values (trend strength
and population variability) of the population model to see how this
altered \(T_{min}\).

In line with past work, we found the \(T_{min}\) increases (i.e.~more
time is required) with decreases in trend strength or autocorrelation
and with increases in variability (citation).

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal
autocorrelation.\label{fig:correlates}}
\end{figure}

@@

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.6 (\(\sigma\)=8.7), with a
wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
expected based on theoretical results (cite papers), the minimum time
series length required is strongly correlated with trend strength,
population variance, and autocorrelation (Fig \ref{fig:correlates}).
Using a generalized linear model, with a Poisson error structure, all
three of these explanatory variables were significant and had large
effect sizes (cite model from supp mat here). Combined trend strength,
population variance, and autocorrelation account for about 70.8 percent
of the variance in minimum time series length required (as proposed by
CITE Zuur).

@@

\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}

@@

@@

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). These simulations support past work\ldots{}

@@

>>>>

Addresses:\emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}
\vspace{3 mm}

@@

Observational studies and population censuses have become a cornerstone
of modern ecological research and conservation biology (Magurran et al.
2010). Longterm data is necessary to both understand population dynamics
and to assess extinction risk of species. Even though many time series
may now be considered ``longterm'', most are still short (i.e.~less than
X years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time
(citations). However, there are of course notable exceptions including
the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others.

Naturally, one might ask, how long of a time series is actually
necessary? This question has important scientific and management
implications. Scientifically, we need to know how long a time series is
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when a trend in abundance over time is actually meaningful or
not. For example, the IUCN Red List Categories and Criteria suggest
under Criterian A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or 3 generations (IUCN 2012).
For both scientific and management questions, because sampling is
typically expensive, we do not want to sample for longer than is
neccessary (Gerber, DeMaster, and Kareiva 1999).

@@

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time. Rhodes and
Jonzen (2011) examined the optimal allocation of effort between spatial
and temporal replicates. Using simple populations models, they found
that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
et al. (2015) also used a modeling approach, but parameterized it for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organims, that 15 years was adequate to assess longterm
trends in abundance. However, these studies, and other past work
(citation), have focused either on theoretical aspects of monitoring
design or focused on single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. In the main manuscript, we determine the minimum time
series length to determine if a population is incresing or decreasing
according to a significant slope coefficient in simple linear
regression.

@@

In figure \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (\(t=40\)). It is
clear that statistical power increases quickly with increases in length
of time sampled (Fig. \ref{fig:theoretical_approach}b). Once power is
greater than XX (the dotted line), that is the minimum time required
(\(t_{min}\)). We can also examine how the minimum time required changes
with effect size and variability (\textbf{make graph of this here})

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, they are a representative sample. We first
take all possible contiguous subsamples of each time series. For
example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continueing until 1 possible contiguous subsample of length 40. Then
we determined the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the longterm, or ``true``, time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1- \beta\), where \(\beta\) is the probability
of a type II error). Lastly, we determine which subsample length is
required to guarantee we have above a certain threshold of statistical
power, which we will assume is 0.8 by convention (citation). A power of
0.8 would imply that given a trend exists, there is a 0.8 probability of
detecting the trend. Complementary, there would be a 0.2 probability of
a false negative.

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{(a) Population size of ANIMAL NAME over time. The red line is
the best fit line from linear regresssion. (b) statistical power for
different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

@@

@@

We constructed a general population model where the trend over time
could be a model parameter. We chose a simple model, but any other
population model could be used. The specific model choice should be
tailored to the population of interest. We then ran simulated time
series of different lengths. From these simulations we determined the
minimum time series length required to achieve a certain level of power
(which we set to 0.8). We were also able to alter the parameter values
(trend strength and population variability) of the population model to
see how this altered \(T_{min}\).

In line with past work, we found the \(T_{min}\) increases (i.e.~more
time is required) with decreases in trend strength and with increases in
variability (citation).

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal
autocorrelation.\label{fig:correlates}}
\end{figure}

@@

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.6 (\(\sigma\)=8.7), with a
wide distribution (Fig. \ref{fig:min_time_dist}). In addition, as
expected based on theoretical results (cite papers), the minimum time
series length required is strongly correlated with trend strength,
population variance, and autocorrelation (Fig \ref{fig:correlates}).
Using a generalized linear model, with a Poisson error structure, all
three of these explanatory variables were significant and had large
effect sizes (see Fig. SX). Combined trend strength, population
variance, and autocorrelation account for about 70.8 of the explained
deviance, which is similar to \(R^2\) (Zuur et al. 2009), in minimum
time series length required.

@@

@@

\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}

@@

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). Our simulations support past work\ldots{}

@@

\hypertarget{ref-Zuur2009}{}
Zuur, Alain F., Elena N. Ieno, Neil J. Walker, Anatoly A. Saveliev, and
Graham M. Smith. 2009. \emph{Mixed Effects Models and Extensions in
Ecology with R}. New York: Springer.

### 2017-06-13 writing/supp_mat_draft.Rmd
moved results from ms to supp mat and expanded supp mat text

1. Detailed example of subsampling and power calculations
2. Figures and results from other questions of interest
3. Additional figures?
4. Additional references

@@

@@

We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text. We then obtain a distribution of the minimum time required to measure the ``true" longterm growth rate (Fig. \reg{fig:min_time_growth_dist}). We see a bimodal distribution with many populations required 30+ years to estimate the longterm growth rate. The large number of short year required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.

```{r, echo=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
# Growth rates calculations for each species
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load('cleaned-data/cleaned_timeseries_database.Rdata')

@@

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
# show gam results for minimium time required
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load('cleaned-data/cleaned_timeseries_database.Rdata')

>>>>

1. Detailed example of subsampling and power calculations
2. Additional figures
3. Figures and results from other questions of interest
4. Additional references

@@

## Additional results from the main manuscript

```{r, echo=F}
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```

In the main manuscript, we examined the minimum time required to detect a significant trend in abundance over time using linear regression. As detailed in the main manuscript the minimum time required was around 15, but there was a wide distribution. Therefore, we were interested in potential explanatory variables of the minimum time required. In the main manuscript, we examined characteristics of the time series itself, like variability, autocorrelation, and the trend in abundance over time. Here, we combined our time series database with a database on life history characteristics of amniotes from (cite other database here). There was life history information available for `r nrow(LPI_pop_info)` populations representing `r length(table(LPI_pop_info$Binomial))` different species, all of which were in the Aves class.

We then correlated minimum time required for each population with its corresponding life history characteristics. In figure \ref{fig:biological_correlates} we examined miminmum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days). None of these variables had much explanatory power in accounting for the variance in the minimum time required. see table \ref{table:model_output}}

```{r, echo=F, fig.cap='Mimimum time required versus (a) generation length (years), (b) litter size (n), (c) adult body mass (grams), (d) maximum longevity (years), (e) egg mass (grams), and (f) incubation (days). The lines in each plot represent the best fit line from linear regression.\\label{fig:biological_correlates}',fig.height=6}

par(mfrow=c(3,2),oma=c(5,4,1,1),mar=c(5,3,0,0))

n=1
x_axis_name_list = c('generation length (years)','litter size (n)','adult body mass (grams)','maximum longevity (years)','egg mass (grams)','incubation (days)')
for (x_axis_name in names(LPI_pop_info)[20:25]){
  plot(LPI_pop_info[[x_axis_name]],LPI_pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info[[x_axis_name]]),lwd=3)

  mtext(text = x_axis_name_list[n],side = 1,line = 3,cex = 1.2)
  mtext(text = paste('(',letters[n],')',sep=''),side = 3,line=-1.5,adj = 0.98,cex=1.2)
  n=n+1
}

mtext(text = 'minimum time required',side = 2,line = 2,cex = 1.4,outer = TRUE)

\pagebreak

In the main text, we explained how the minimum time required strongly correlated with the trend strength, temporal autocorrelation, and variance in population size. Here we use a generalized linear model framework with a Poisson error structure to determine drivers of the minimum time required. In figure, \ref{fig:poisson_model} we show a set of residual plots for the regression. We then show the coefficient estimates and levels of significance in table \ref{table:model_output}.

```{r poisson_model, echo=F,include=T,fig.cap='Output of poisson regression...\\label{fig:poisson_model}'}
setwd("~/Desktop/Research/time-series-project")
load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

# GLM with Poisson error structure (include all terms - it is best not to include abs_overall_trend with variance as they are strongly corrrelated to one another)
pop_info$abs_overall_trend = abs(pop_info$overall_trend)
poisson_model <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)

  par(mfrow=c(2,2),mar=c(5,5,1,2),oma=c(0,0,0.5,0))
    plot(poisson_model,cex.lab=1.2,las=1,cex.axis=1.2,main=' ',pch=16, col=rgb(0.5,0.5,0.5,0.4))
#R^2 calculation (pg. 218 Zuur et al. 2009 mixed effects book)
R2_fullmodel <- 100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance)

poisson_model_no_variance <-glm(min_time_for_power~ abs_overall_trend + autocorrelation + gen_len,family = 'poisson',data=pop_info)
R2_no_variance <- 100*(poisson_model_no_variance$null.deviance - poisson_model_no_variance$deviance)/(poisson_model_no_variance$null.deviance)

poisson_model_no_trend <-glm(min_time_for_power~ autocorrelation + variance + gen_len,family = 'poisson',data=pop_info)
R2_no_trend <- 100*(poisson_model_no_trend$null.deviance - poisson_model_no_trend$deviance)/(poisson_model_no_trend$null.deviance)

```{r,echo=F,message=F,warning=F}

require(knitr)
knitr::kable(summary(poisson_model)$coef,caption = 'Stuff here\\label{table:model_output}')

Our model with trend strength, autocorrelation, variance, and generation length accounted for `r round(R2_fullmodel,2)`% of the variation in the minimum time required. However, we also found trend strength and variance to be strongly correlated with one another. Therefore, we ran two additional models with either trend strength or variance, but not both together. This resulted in lower explained deviance (analogue to $R^2$) of `r round(R2_no_variance,2)`% and `r round(R2_no_trend,2)`%, respectively.

\pagebreak

@@

We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text. We then obtain a distribution of the minimum time required to measure the ``true" longterm growth rate (Fig. \ref{fig:min_time_growth_dist}). We see a bimodal distribution with many populations required 30+ years to estimate the longterm growth rate. The large number of short year required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.

```{r, echo=F,eval=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
# Growth rates calculations for each species
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load('cleaned-data/cleaned_timeseries_database.Rdata')

@@

```{r,echo=F,eval=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
# show gam results for minimium time required
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load('cleaned-data/cleaned_timeseries_database.Rdata')

### 2017-06-13 writing/supp_mat_draft.tex
moved results from ms to supp mat and expanded supp mat text

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Detailed example of subsampling and power calculations
\item
  Figures and results from other questions of interest
\item
  Additional figures?
\item
  Additional references
\end{enumerate}

@@

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) abundance over time, (b) mean and standard deviation of
growth rate for subsamples of entire time series, and (c) the percent
error between mean estimated growth rate and the true longterm growth
rate. The vertical bar denotes the minimum time required to estimate
growth rate within 10\% error.\label{fig:growth_rate}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Histogram of the minimum time required in order to estimate the
longterm growth rate within 20\% error.\label{fig:min_time_dist}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{results here from GAM.\label{fig:gam_example}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{results here from GAM.\label{fig:gam_result}}
\end{figure}

@@

\subsubsection{References}\label{references}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Detailed example of subsampling and power calculations
\item
  Additional figures
\item
  Figures and results from other questions of interest
\item
  Additional references
\end{enumerate}

@@

\subsection{Additional results from the main
manuscript}\label{additional-results-from-the-main-manuscript}

In the main manuscript, we examined the minimum time required to detect
a significant trend in abundance over time using linear regression. As
detailed in the main manuscript the minimum time required was around 15,
but there was a wide distribution. Therefore, we were interested in
potential explanatory variables of the minimum time required. In the
main manuscript, we examined characteristics of the time series itself,
like variability, autocorrelation, and the trend in abundance over time.
Here, we combined our time series database with a database on life
history characteristics of amniotes from (cite other database here).
There was life history information available for 547 populations
representing 315 different species, all of which were in the Aves class.

We then correlated minimum time required for each population with its
corresponding life history characteristics. In figure
\ref{fig:biological_correlates} we examined miminmum time required
versus generation length (years), litter size (n), adult body mass
(grams), maximum longevity (years), egg mass (grams), and incubation
(days). None of these variables had much explanatory power in accounting
for the variance in the minimum time required. see table
\ref{table:model_output}\}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

\pagebreak

In the main text, we explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
variance in population size. Here we use a generalized linear model
framework with a Poisson error structure to determine drivers of the
minimum time required. In figure, \ref{fig:poisson_model} we show a set
of residual plots for the regression. We then show the coefficient
estimates and levels of significance in table \ref{table:model_output}.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
\caption{Output of poisson regression\ldots{}\label{fig:poisson_model}}
\end{figure}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Stuff here\label{table:model_output}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 3.7141297 & 0.0274350 & 135.379203 &
0.0000000\tabularnewline
abs\_overall\_trend & -104.4974288 & 3.0997732 & -33.711314 &
0.0000000\tabularnewline
autocorrelation & -0.4905888 & 0.0431062 & -11.380936 &
0.0000000\tabularnewline
variance & 18.7756472 & 0.6409982 & 29.291262 & 0.0000000\tabularnewline
gen\_len & 0.0053584 & 0.0025995 & 2.061352 & 0.0392695\tabularnewline
\bottomrule
\end{longtable}

Our model with trend strength, autocorrelation, variance, and generation
length accounted for 72.55\% of the variation in the minimum time
required. However, we also found trend strength and variance to be
strongly correlated with one another. Therefore, we ran two additional
models with either trend strength or variance, but not both together.
This resulted in lower explained deviance (analogue to \(R^2\)) of
53.08\% and 45.04\%, respectively.

\pagebreak

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) abundance over time, (b) mean and standard deviation of
growth rate for subsamples of entire time series, and (c) the percent
error between mean estimated growth rate and the true longterm growth
rate. The vertical bar denotes the minimum time required to estimate
growth rate within 10\% error.\label{fig:growth_rate}}
\end{figure}

We applied the same calculations to 1032 populations, as we did in the
main text. We then obtain a distribution of the minimum time required to
measure the ``true" longterm growth rate (Fig.
\ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
populations required 30+ years to estimate the longterm growth rate. The
large number of short year required is due to cases where the entire
time series is consistently increasing or decreasing at the same rate.

\clearpage

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{results here from GAM.\label{fig:gam_example}}
\end{figure}

@@

@@

\clearpage
\#\#\# References
